Ablation study on the residual adapter in Neural Machine Translation

### 2020-08-14 2020/wmt_20/emnlp2020.tex
local changes

\begin{document}
\maketitle
\begin{abstract}
Among the approaches for adapting a pretrained NMT model to a specific domain, finetuning is the dominant method. Finetuning proposes 2 approaches: simply continuing training all the parameters \cite{Luong15stanford}; training addition adapters while freezing the pretrained model\cite{Bapna19simple,Vilar18learning}. The second approach has several advantages including preserving the pretrained model and the legerity of the adapters. However, the behavior of these adapters has not been further studied. The objective of this paper is to give an ablation study on the use of residual adapter in the domain adaptation problem.
\mpTodo{correcting abstract}

@@

\end{abstract}
\section{Introduction } \label{sec:intro}
\mpTodo{write introduction} \fyDone{Citations in chronological order}\fyDone{Split long sentences}
Owing to multiple improvements, Neural Machine Translation(NMT) \cite{Kalchbrenner13recurrent,Sutskever14sequence,Bahdanau15learning,Vaswani17attention} nowadays delivers useful outputs for many language pairs. However, as many deep learning model, NMT models need to be trained with sufficiently large amounts of data to reach their best performance. Therefore, the quality of translation of NMT model is still limited in low-resource language and low-resourced domain conditions \cite{duh13adaptation,zoph16transfer,koehn17six}. While many approaches have been proposed to improve the quality of NMT models in low-resource domains (see the recent survey of \citet{Chu18asurvey}), full fine-tuning of a generic baseline model remains the dominant supervised approach when adapting NMT models to specific domains \cite{Luong15stanford,neubig18rapid}.

Under this view, building adapted systems is a two-step process: (a) one first trains NMT with the largest possible parallel corpora, possibility aggregating texts from multiple, heterogeneous sources; (b) assuming that in-domain parallel documents are available for the domain of interest, one then adapts the pre-trained model by resuming training with the sole in-domain corpus. It is a conjecture that pre-trained model constitute a better initialization than a random one, especially when the adaptation data is scarce. Indeed, studies of transfer learning for NMT such as \cite{artetxe20cross,aji20neural} have confirmed this claim by extensive experiments. Full fine-tuning, that adapts all the parameters of a baseline model usually significantly improves the quality of the NMT for the chosen domain. However, it also yields large losses in translation quality for other domains, a phenomenon referred to as ``catastrophic forgetting'' in the neural network literature \cite{McCloskey89catastrophic}. Therefore, a fully fine-tuned model is only useful to one target domain. As the number of domains to handle grows, training and maintaining a separate model for each task can quickly becomes tedious and resource-expensive .\fyDone{Fix this sentence.}

\cite{Vilar18learning,Bapna19simple} have recently proposed a simple and lightweight domain adaptation method that also preserves the value of pre-trained models, based on small adapters component that are plugged in each hidden layer. These adapters are trained only with the in-domain data, keeping the pre-trained model frozen. Because these additional adapters are very small compared to the size of the baseline model, their use significantly reduces the cost of training and maintaining fine-tuned models, while still delivering a performance that is close to that of full fine-tuning.

In this paper, we would like to extend this architecture to improve NMT in several settings that still challenge automatic translation, such as translating texts from multiple topics, genre or domains, in the face of unbalanced data distribution. Furthermore, as the notion of ``domains''  is not always well established, another practical setting is the translation of texts mixing several topics / domains. Finally, another requirement is the need to also translate texts from domain unseen in training, based only on the unadapted system, which should be made as strong as possible. 
% Residual adapters allow us to adapt NMT model to any specific domain in a computationally cheap way. Could we adapt NMT model to noisy text and topical text at once? While residual adapters are good at adapting to one specific domain, their performance still dramatically decrease for the other domains seen in training, as we show in our experiments. Therefore, we have to decide whether to use residual adapters manually, i.e, we have to know to which domain the text belongs a priori. Therefore, we would like to fuse a domain classifier to the architecture in order to weight the contribution of the adapters with respect to the topic-relatedness of the text.
\fyDone{Say differently: various implementations}

In this context, our main contribution is a thorough experimental study of the use of residual adapters in a multi-domain translation adaptation. We notably explore ways to adjust and/or regularize adapter modules to handle situations where the adaptation data is very small. We also propose and contrast two new variants of the residual architecture: in the first one (highway residual adapters), adaptation still affects each layer of the architecture, but its effect is delayed till the last layer, thus making the architecture more modular and adaptive; in the second one (gated residual adapters), we explore ways to improve the performance in the face of train-test data mismatch. We experiment with two langage pair and report results that demonstrate the flexibility and effectiveness of these architectures. 
\fyDone{One bit of a conclusion here}\fyFuture{Build a proper training scenario for these two conditions}.

@@

\subsection{Ablation study on position and number of residual adapters \label{ssec:ablatation}}
\fyTodo{Move this}
In this section, we would like to analyze the impact of position and number of residual adapters involved in the adapted model. In fact, the importance of hidden layers varies with respect to their position in the model. It is conjectured that the higher layers represent more global patterns such as semantic while the lower layers represent more local patterns such as syntactic\fyTodo{Missing reference here}. The domain shift in local patterns and global patterns has not yet studied. In this paper, we do not intend to study this aspect of domain adaptation problem. We would like to give an extensive comparison between domain adaptation in different levels in NMT model.\fyTodo{Fix style here} Because the set of possible configurations is large, we only perform experiments on layers at position: 2, 4, 6 which covers a wide range in the hierarchy of attention layers.

Second, we would like to observe how much the number of residual adapters affects the performance of the adapted model. By the limit of computation resource, we limit our study to the case of 3 adapters at positions: 2,4,6. By doing this, we could cover 3 situations: 1 adapter (mentioned in the previous paragraph), 3 adapters and 6 adapters (in the normal setting) in the encoder or the decoder.\fyTodo{We have 12 layers, have we not ?}

@@

\subsection{Highway ...}
Finally, we would like to assess the performance of a variant of residual adapters in which the adapters go directly to the end of the stack of attention layers. We graphically demonstrate this version in Figure~\ref{fig:hrl-architecture}. The benefit of this variant is that we can manipulate all values of the adapters at once.

\subsection{Word-Level Adaptive Domain Adaptation \label{ssec:wada}}
\mpTodo{Formalizing problem, network design, training algorithm}
Residual adapters allow us to adapt NMT models to a new domain without altering the pretrained parameters. However, we still have to decide in which domain a sentence is translated to chose a suitable adapter.\fyTodo{Lack of context - domain errors} Therefore, we face to another problem which is the error of domain prediction. Choosing wrong adapter to translate a sentence can be dramatically bad because a specialized model usually has bad performance for out-of-domain instances \cite{McCloskey89catastrophic}. However, we still have another option that is to use the pretrained generic model without adaptation. Therefore, we could aim to achieve a performance at least as good as the generic model. We would like to present an adaptive version of residual adapter below.
\fyTodo{Consistency of notations wrt section 2.1}
We consider a generic encoder-decoder architecture for a multi-domain sequence to sequence learning. Let $h \in \R^d$ the output of the encoder, \cite{Bapna19simple} proposes to use an adapted representation for domain $k$ defined as $h' = h + a_k(h)$. This means that all words in all sentences in domain $k$ will use the adapter module represented by $a_k$. In a multilayer version with highway connections (see Figure~\ref{fig:hrl-architecture}), $a_k(h) = \sum_{l} a_{kl}(h^l)$ where $h^l$ is the distributed representation of the sequence at the $l^{th}$ layer.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{fig/highway_residual}
  \caption{Highway multi-domain network with residual layers}
\label{fig:hrl-architecture}
\end{figure}

@@

Early approaches to the multi-domain setting were proposed for statistical MT, either considering multiple data sources (eg.\ \cite{Banerjee10combining,Clark12onesystem,Sennrich13multidomain,Huck15mixeddomain}) or domains containing several topics \cite{Eidelman12topic,Hasler14dynamic-topic}. Two main strategies are considered: feature-based methods, where domain/topic labels are integratetd through supplementary feature and instance-based methods, involving a measure of similarity between train and test domains.

The former approach has also been adapted to NMT: \newcite{Kobus17domaincontrol,Tars18multidomain} use an additional domain feature in an RNN model, in the form of an extra domain-token or of additional domain-features associated with each word. \citet{Chen16guided} apply domain control on the \emph{target} side, using a topic vector to describe the whole document context. Similar ideas are developed in \cite{Chu18multilingual,Pham19generic}, where domain differences and similarities are enforced through parameter sharing schemes. Parameter-sharing also lies at the core of the work by \citet{Jiang19multidomain}, who consider a Transformer model containing both domain-specific heads and domain-agnostic heads.

\citet{Britz17mixing} studies three general techniques to take domain information into account in training: they rely on either domain classification or domain normalization on the source or target side. One contribution of this study is the proposal of an adversarial training scheme to normalize representations across domains and make the combination of multiple data sources more effective. Similar techniques (parameter sharing, automatic domain classification / normalization) are at play in \cite{Zeng18multidomain}: in this work, the lower layers of the MT use auxiliary classification tasks to disentangle domain specific from domain-agnostic representations. These two representations are then processed separately, then merged to compute the final translation.

\citet{Farajian17multidomain,li-etal-2018-one} are two recent representatives of the instance-based approach: for each test sentence, a small adaptation corpus is collected based on similarity measures, and used to fine-tune a mix-domain model.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Among the approaches for adapting a pretrained NMT model to a specific domain, finetuning is the dominant method. Finetuning proposes 2 approaches: simply continuing training all the parameters \cite{Luong15stanford}; training addition adapters while freezing the pretrained model\cite{Bapna19simple, Vilar18learning}. The second approach has several advantages including preserving the pretrained model and the legerity of the adapters. However, the behavior of these adapters has not been further studied. The objective of this paper is to give an ablation study on the use of residual adapter in the domain adaptation problem.
\mpTodo{correcting abstract}

@@

\end{abstract}
\section{Introduction } \label{sec:intro}
\mpTodo{write introduction} \fyTodo{Citations in chronological order}\fyTodo{Split long sentences}
Owing to multiple architectural improvements, Neural Machine Translation(NMT) \cite{Kalchbrenner13recurrent,Sutskever14sequence,Bahdanau15learning,Vaswani17attention} nowadays delivers useful outputs for many language pairs. However, as many deep learning model, NMT models need to be trained with sufficiently large amounts of data to reach their best performance. Therefore, the quality of translation of NMT model is still limited in low-resource language and low-resourced domain conditions \cite{duh13adaptation, zoph16transfer,koehn17six}. While many approaches have been proposed to improve the quality of NMT models in low-resource domains (see the recent survey of \citet{Chu18asurvey}), full fine-tuning of a generic baseline model remains the dominant supervised approach when adapting NMT models to specific domains \cite{Luong15stanford,neubig18rapid}.

Under this view, building adapted systems is a two-step process: (a) one first trains NMT with the largest possible parallel corpora, possibility aggregating texts from multiple, heterogeneous sources; (b) assuming that in-domain parallel documents are available for the domain of interest, one then adapts the pretrained model by resuming training with the sole in-domain corpus. It is a conjecture that pretrained model constitute a better initialization a the random initialization, especially when the adaptation data is scarce. Indeed, studies in transfer learning for NMT such as \cite{artetxe20cross,aji20neural} have confirmed this claim by extensive experiments. Full fine-tuning, that is adapting all the parameters of a baseline model usally significantly improves the quality of the NMT for the chosen domain. However, it also implies large losses in translation quality for other domains, a phenomenon referred to as ``catastrophic forgetting'' in the neural network literature \cite{McCloskey89catastrophic}. Therefore, a fully fine-tuned model is only useful to one target domain. As the number of domains to handle grows, training and maintaining such models can consume a lot of resource.\fyTodo{Fix this sentence}

\cite{Vilar18learning,Bapna19simple} have recently proposed a simple and lightweight method to preserve the value of pretrained models by plugging small adapters to hidden layers. These adapters are trained only with the in-domain data, keeping the pretrained model frozen. Because these additional adapters are very small compared to the size of the baseline model, their use significantly reduces the cost of training and maintaining fine-tuned models, while still delivering a performance that is on par with full fine-tuning.

In this paper, we would like to extend this architecture to improve NMT in several specific situations that still challenge automatic translation, such as translating noisy inputs or translating topical texts\fyTodo{multidomain, multitopic, unbalance data}. Residual adapters allow us to adapt NMT model to any specific domain in a computationally economical way. Could we adapt NMT model to noisy text and topical text at once? While residual adapters are good at adapting to one specific domain, their performance still dramatically decrease for the other domains seen in training, as we show in our experiments. Therefore, we have to decide whether to use residual adapters manually, i.e, we have to know to which domain the text belongs a priori. Therefore, we would like to fuse a domain classifier to the architecture in order to weight the contribution of the adapters with respect to the topic-relatedness of the text.\fyTodo{Say differently: various implementations}

Our contribution is an thorough experimental study on the use of residual adapters in multidomain adapation. We notably try to answer the following three questions:
\begin{itemize}
\item How the number and position of adapters affect the performance.
\item How to train the residual adapters.
\item How to render the combination of NMT model and residual adapters more robust to out-of-domain examples.
\end{itemize}
\fyTodo{One bit of a conclusion here}

@@

\subsection{Ablation study on position and number of residual adapters \label{ssec:ablatation}}
\fyTodo{Move this}
In this section, we would like to analyze the impact of position and number of residual adapters involved in the adapted model. In fact, the importance of hidden layers varies with respect to their position in the model. It is conjectured that the higher layers represent more global patterns such as semantic while the lower layers represent more local patterns such as syntactic \cite{raganato18analysis}\fyTodo{Missing reference here}. The domain shift in local patterns and global patterns has not yet studied. In this paper, we do not intend to study this aspect of domain adaptation problem. We would like to give an extensive comparison between domain adaptation in different levels in NMT model.\fyTodo{Fix style here} Because the set of possible configurations is large, we only perform experiments on layers at position: 2, 4, 6 in each side of NMT model (encoder and decoder) which covers a wide range in the hierarchy of attention layers.

Second, we would like to observe how much the number of residual adapters affects the performance of the adapted model. By the limit of computation resource, we limit our study to the case of 3 adapters at positions: 2,4,6 in each side of the NMT model. By doing this, we could cover 3 situations: 1 adapter (mentioned in the previous paragraph), 3 adapters and 6 adapters (in the normal setting) in each side (encoder and decoder).\fyTodo{We have 12 layers, have we not ?}

@@

\subsection{Highway connection}
Finally, we would like to assess the performance of a variant of residual adapters in which the adapters go directly to the end of the stack of layers. We graphically demonstrate this version in Figure~\ref{fig:hrl-architecture}. In this version, only the last layers of the encoder and the decoder are modified.

\begin{equation}
\bar{h}^n = h^n + \displaystyle{\mathop{\sum}_{1\leq i\leq n}} ADAP^i(h^i)
\end{equation}

The benefit of this variant is that we can manipulate all values of the adapters at once.

\subsection{Word-Level Adaptive Domain Adaptation \label{ssec:wada}}
\mpTodo{Formalizing problem, network design, training algorithm}
Residual adapters allow us to adapt NMT models to a new domain without altering the pretrained parameters. However, we still have to decide in which domain a sentence is translated to chose a suitable adapter.\fyTodo{Lack of context - domain errors} Therefore, we face to another problem which is the error of domain prediction. Choosing wrong adapter to translate a sentence can be dramatically bad because a specialized model usually has bad performance for out-of-domain instances \cite{McCloskey89catastrophic}. However, we still have another option that is to use the pretrained generic model without adaptation. Therefore, we could aim to achieve a performance at least as good as the generic model. We would like to present an adaptive version of residual adapter below.
\fyTodo{Consistency of notations wrt section 2.1}
We consider a generic encoder-decoder architecture for a multi-domain sequence to sequence learning. Let $h \in \R^d$ the output of the encoder, \cite{bapna19simple} proposes to use an adapted representation for domain $k$ defined as $h' = h + a_k(h)$. This means that all words in all sentences in domain $k$ will use the adapter module represented by $a_k$. In a multilayer version with highway connections (see Figure~\ref{fig:hrl-architecture}), $a_k(h) = \sum_{l} a_{kl}(h^l)$ where $h^l$ is the distributed representation of the sequence at the $l^{th}$ layer.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{fig/highway_residual}
  \caption{Highway multi-domain network with residual layers}
\label{fig:hrl-architecture}
\end{figure}

@@

%; scores are reported separately per domain, suggesting that all domains are equally important at test time.
% Adversarial adpatation is used in many domains
% for QA @misc{lee2019domainagnostic,
%    title={Domain-agnostic Question-Answering with Adversarial Training},
%    author={Seanie Lee and Donggyu Kim and Jangwon Park},
%    year={2019},
%    eprint={1910.09342},
%    archivePrefix={arXiv},
%    primaryClass={cs.CL}
%}, NLU and many others

Early approaches to the multi-domain scenario were proposed for statistical MT, either considering multiple datansources (eg.\ \cite{Banerjee10combining,Clark12onesystem,Sennrich13multidomain,Huck15mixeddomain}), or domains containing of several topics \cite{Eidelman12topic,Hasler14dynamic-topic}. Two main strategies are considered: feature-based methods, where domain/topic labels are integratetd through supplementary feature and instance-based methods, involving a measure of similarities between train and test domains.

The former approach has also been adapted to NMT: \newcite{Kobus17domaincontrol,Tars18multidomain} use an additional domain feature in an RNN model, either in the form of an extra (initial) domain-token or in the form of additional domain-feature associated to each word. \citet{Chen16guided} apply domain control on the \emph{target} side, using a topic vector to describe the whole document context. Similar ideas are developed in \cite{Chu18multilingual,Pham19generic}, where domain differences and commonalties are enforced through parameter sharing schemes.
Parameter-sharing is also used by \citet{Jiang19multidomain}, who consider a Transformer model containing both domain-specific heads and domain-agnostic heads.

\citet{Britz17mixing} studies three general techniques to take domain information into account in training: the rely on either domain classification (or domain normalization) on the source or target side. One contribution of this study is the use of adversarial training to normalize representations and make the combination of multiple data sources more effective. Similar techniques (parameter sharing, automatic domain classification / normalization) are at play in \cite{Zeng18multidomain}: in this work, the lower layers of the MT use auxiliary classification tasks to disentangle domain specific from domain-agnostic representations. These two representations are then processed separately, then merged to compute the final translation.

\citet{Farajian17multidomain,li-etal-2018-one} are two recent representatives of the instance-based approach: for each test sentence, a small adaptation corpus is collected based on similarity measures, and used to fine-tune a mix domain model.
